---
title: "Lab 3 : Graphing with ggplot"
output:
  html_document:
    toc: true
    toc_depth: 4
    theme: united
    highlight: tango
editor_options: 
  chunk_output_type: console
---

## Learning objectives

* ggplot

## Introduction to R Graphics

R provides comprehensive graphics utilities for visualizing and exploring scientific data. To date we have been making a few plots using the R Base Graphics. In addition, several more recent graphics environments extend these utilities. These include the `grid`, `lattice` and `ggplot2` packages. The `ggplot2` environment is by far the most popular and used for many R packages and in many scientific publications. 

### Grammar of Graphics

`ggplot2` is meant to be an implementation of the Grammar of Graphics, hence the gg in ggplot. The basic notion is that there is a grammar to the composition of graphical components in statistical graphics. By direcly controlling that grammar, you can generate a large set of carefully constructed graphics from a relatively small set of operations. As Hadley Wickham (2010), the author of `ggplot2` said,

"A good grammar will allow us to gain insight into the composition of complicated graphics, and reveal unexpected connections between seemingly different graphics.

### Installation

`ggplot2` is a R package, which is part of the `tidyverse` collection which is already instlled. You can load just `ggplot`

```{r}
library(ggplot2)
```

or all `tidyverse` which loads `ggplot2` with other packages

```{r}
library(tidyverse)
```

### Tutorials and resources

You can make amazing graphs with ggplot, but there is a long learning curve so we will have multiple lab sessions on ggplot and graphing. Here are a few different resources for `ggplot`.

* Hadley Wickham and Garrett Grolemund released a new book in 2017 [R for Data Science](http://r4ds.had.co.nz/).Chapter 3.1 to 3.5 covers material for this week's lab in ggplot.
* Another resource good for those with a visual learning style is [Maria Nattestad's Youtube videos](https://www.youtube.com/channel/UC2bWYX9h1KlaGWfTDuhASWg)

### Key board shortcuts

* Selecting `shift` `ctrl` `m` at the same time is a keyboard shortcut for `%
* Selecting `ctr` `alt` `i` at the same time is a keyboard shortcut for a new r code 
To see more shortcuts in the menu `Tools` → `Keyboard Shortcuts Help`.

## On the Computer

Today's lab is a walk through of Data Carpentry's [ggplot tutorial](https://datacarpentry.org/R-ecology-lesson/04-visualization-ggplot2.html) up to (but not including) the sections on `Arranging and Explorting plots`.

## What to upload to Moodle

Create a report using R Markdown with all the [ggplot tutorial](https://datacarpentry.org/R-ecology-lesson/04-visualization-ggplot2.html) examples and with the challenges listed at the end.

